Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-13 Manuscript/manuscript.tex

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

Zwitterionic phosphocholine (PC) lipid bilayers are commonly used model systems for
cellular membranes. Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in PC lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Interpretation most of these experiments suggests that Na$^+$ ions exhibit
negligible binding in PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules. Atomistic resolution
molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. This study presented 
a direct comparison between ion binding affinity between simulations and
experiments by using NMR data for lipid headgroup order 
parameters and the electrometer concept~\cite{seelig87}.
Using massive amounts of data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of the lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased back
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in the Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules.
  JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer
  charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever)
  and ``semi-scaled'' models.}
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stronger electrostatic interactions with them.
After reducing the $\sigma$ parameters by a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
back in agreement with the experimental value (see Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding between experiments and simulations was compared by 
using lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid,  $X^\pm$~\cite{seelig87}.
Change in the order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

The change of the head group order parameters is empirically quantified
as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because recent analysis concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. This is rationalized as a change
of lipid head group dipole tilt toward water phase with bound positive
charge and {\it vice versa} with negative charge \cite{seelig87}.

The concept can be used to compare the ion binding affinity in lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

The measured order parameter change depends on the response of 
the headgroup on the bound charge, i.e. $m_i$ in Eq.~\ref{OPchangeEQ},
and the ion binding affinity. Thus, the former property has to be
well quantified before using the electrometer concept to
analyze binding affinities. This is done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response also in simulations,
we use experimental data for dihexadecyldimethylammonium (2C$_{16}$$^+$N2C$_1$Br$^-$)
in a POPC bilayer \cite{scherer89}. Dihexadecyldimethylammonium
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems \cite{scherer87}.

\subsection{Salt concentrations and binding affinity}
Experimental studies providing the lipid head group order
parameters report salt concentrations in two different ways.
For DPPC, salt concentrations before solvating
the lipids are used~\cite{akutsu81}. Alternatively, for POPC,
the salt concentration of supernatant after solvation is
measured using atomic absorption~\cite{altenbach84}.
\todo{JOE: I'm afraid that the above description might sound to the reader
  like for some reasons one method is used with DPPC, but a different one with POPC.
  I think simply writing one way is\dots another way is \dots (or similarly) will be less confusing.}
In this work, we will use the latter definition. For this we
will calculate the salt concentration in the aqueous bulk region,
\textit{i.e.}, the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the concentration was calculated as in the DPPC experiments.
Despite of the measurable differences between these two concentrations for CaCl$_2$ systems,
the qualitative conclusions are affected neither in this
nor in our previous work~\cite{catte16}.

To quantify the ion binding affinity to a membrane, we
calculated the relative surface excess of ions with respect to water $\Gamma_i^w$  \cite{chattorajBOOK}.
This quantity can be used to compare binding affinity between different models,
because it does not depend on the definition of the interfacial region
or the position of the Gibbs dividing plane between two bulk regions.
\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
Therefore, we consider the hydrophobic interior of a membrane having zero concentrations of ions and water
as a bulk region, which is separated by a monolayer interface from 
a another region with bulk concentration of ions in water.
Since the boundary of the latter region is set to the edge of simulation box,
all molecules in simulation are within the defined interfacial region. 
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the size of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

Lipid bilayer structure without ions was validated against NMR and 
\mbox{x-ray} scattering experiments by calculating order parameters
for C-H bonds and form factors from our simulations. The former validates the structures sampled by individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

The order parameters were calculated from simulations for all C-H bonds
in lipid molecules by using Eq. \ref{OP}. Form factors were calculated 
from equation
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length in the bulk region,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning until the water bulk region, $D$.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
The simulated systems consisted of POPC bilayer in pure water
or in varying salt concentrations.
In particular, the periodic orthorhombic simulation box contained 
128 POPC molecules and approximately 50~water molecules per each lipid.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
presented in Supporting Information (SI). 
We used the ECC-ions model for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion models by Dang~\cite{smith94,chang1999,dang2006}, and by \AA{}qvist~\cite{aqvist90}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium \\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH2{)}_{15}-CH_3)(-[:30]{(}CH2{)}_{15}-CH_3)
} , 
\vspace{0.5cm} \\
which is one of the cationic surfactants used to experimentally quantify the electrometer concept~\cite{scherer89}.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
%Chloride counter ions were used
%in simulations, because parameters for bromide, which was used in the experiment \cite{scherer89},
%were not available in the standard Gromacs files for Amber force field.
The systems were simulated for 200~ns using Lipid14 model for POPC where reasonable lipid neighbor exchange occurs.
First 20~ns were omitted from the analysis.

The same systems were also simulated with the ECC-lipid model for POPC using the same setup. 
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, i.e., $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: Chemical structure of POPC with definitions of different order parameters calculated.
    Middle: Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  
    from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.
    Error bars are approximately the size of the points for head group order parameters.
    Point size for the order parameters of tails are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    } 
\end{figure}

\begin{table}
  \caption{Area per lipid (APL) from different models of POPC with no ions\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 62.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

%In order to validate the structural properties of the newly developed ECC-lipids model
%we first compared our simulation results without any ions to
%NMR order parameters measurements and x-ray scattering form factors 
%(Fig.~\ref{simVSexpNOions}). 
The x-ray scattering form factor (Fig.~\ref{simVSexpNOions}) and
the area per lipid (Table~\ref{tab:apls}) of POPC bilayer simulated
with the ECC-lipid model are in good agreement with the experimental results,
as well as with the Lipid14 model. Thus, we conclude that the lipid bilayer dimensions
are well captured by the developed ECC-lipid model. 
Areas per lipid of ECC-lipids with various other water models are documented in SI, table~\ref{tab:apls_si}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
used to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. A direct comparison of
ion binding affinities to PC bilayers was presented between simulations and
experiments by using the electrometer concept, which is based on the experimental NMR data
for the lipid headgroup order parameters~\cite{seelig87}.
Using massive amounts of data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

@@

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of the lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased back
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in the Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules. \\
  JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer
  charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever)
  and ``semi-scaled'' models. \\
SAMULI: There is now a paragraph in the conclusions, which mentions this topic.}
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stronger electrostatic interactions with them.
After reducing the $\sigma$ parameters by a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
back in agreement with the experimental value (see Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations by 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid~\cite{seelig87}.
The change of order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

The concept can be used to compare the ion binding affinity in lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{catte16,ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

The relation between bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with bound positive charge and increase with bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt toward water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

The measured order parameter change depends on both, the head group response 
to the bound charge and the amount of bound charge, i.e. $m_i$ and $X^\pm$
in Eq.~\ref{OPchangeEQ}, respectively. The former property has to be
well quantified before the electrometer concept can used to 
analyze the binding affinities, as done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response to the bound charge in simulations,
we use the experimental data for dihexadecyldimethylammoniumbromide %(2C$_{16}$$^+$N2C$_1$Br$^-$)
mixed within a POPC bilayer~\cite{scherer89}. Dihexadecyldimethylammonium\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
The early experimental electrometer concept data for the lipid head group
order parameter changes was reported as a function 
of the salt concentration in water before solvating
the lipids~\cite{akutsu81}. The later study used atomic absorption
spectroscopy and reported the order parameter changes as
a function of the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was calculated from 
the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the ion concentrations were calculated in water before solvating the lipids 
as in the earlier experiments~\cite{akutsu81}.
Despite of the measurable differences between these two
definitions of concentrations for CaCl$_2$ systems,
the qualitative conclusions in this or in the previous work~\cite{catte16} are
not affected by this.

To quantify the ion binding affinity to a lipid bilayer, we
calculated the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
This quantity was chosen because it does not depend on
%the definition of the interfacial region or
the position of the Gibbs dividing plane between two bulk regions.
%\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
%SAMULI: Here is a new trial.
Here we assume that the interface locates between the hydrophobic interior
of a lipid bilayer and the bulk water region far from the membrane.
The bulk concentration of ions and water is zero inside the bilayer.
The concentrations in bulk water region can be calculated from 
the farthest point from both lipid leaflets in the water phase.
The region between these boundaries contains all the ions and water
molecules in the simulation box.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the size of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

The structure of lipid bilayers in simulations without ions were validated against NMR and 
\mbox{x-ray} scattering experiments by calculating the order parameters
for C-H bonds and the scattering form factors. The former validates the structures sampled by individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

The order parameters were calculated from simulations for all C-H bonds
in lipid molecules by using Eq. \ref{OP}. Form factors were calculated 
from equation \todo{As Hector suggested, it might be better to write the simpler
form for this equation.}
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length in the bulk region,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning until the water bulk region, $D$.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
Simulations of POPC bilayer in pure water or in varying salt concentrations
contained 128 POPC molecules and approximately 50~water molecules per each lipid
in the periodic orthorhombic simulation box.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model
\todo{More justification for the choosing the OPC3 water model are needed.
It might be good to show the comparison with the scattering data in bulk water in SI.}. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
\todo{The normal TIP3P was tested as well, right?} presented in Supporting Information (SI). 
The ECC-ion models were used for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulation with the Lipid14 model was also ran with the ion model by Dang et al.~\cite{smith94,chang1999,dang2006}.
Simulation data for the Lipid14 model with \AA{}qvist~\cite{aqvist90} ion model
was taken directly from \cite{lipid14POPC0mMNaClfiles}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The parameters are available at \cite{??} \todo{To be added}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. Chloride counter ions were used
in simulations, because parameters for bromide, the counterion in the experimental data \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
The Lipid14 model was used for POPC.
The first 20~ns of the total simulation time of 200~ns
was considered as an equlibration time and was omitted from the analysis.
A reasonable lipid neighbor exchange occured during the simulation.

The same systems were also simulated with the ECC-lipid model for POPC using the same setup.
\todo{Which water model was used in these simulations?}
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, i.e., $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids. 
\todo{Parameters are available at ??}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: The chemical structure of POPC and the labeling of the carbon segments.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipid models
    compared with the experimental values from \cite{ferreira13}.
    The size of the points for the head group order parameters correspond the
    of the error bars ($\pm 0.02$ for experiments \cite{botan15,ollila16} and ?? for simulations).
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipid models compared with experiments~\cite{Kucerka2011}.
  } 
  \todo{Increase the height of the order parameter figure} \\
  \todo{sn-1 and sn-2 order parameters should be somehow labelled in the order parameter figure. Maybe empty and filled points?} \\
  \todo{Add size of the error bars in simulations to the caption.} \\
  \todo{Would it be possible to increase the size of points for acyl chain order parameters only in z-direction such that
    it would correspond the error bars?} \\
  \todo{x-axis in form factor plot from 0 to 0.6, where experimental data ends. y-axis from 0 to $\sim$2.5 to remove the empty space.} \\
\end{figure}

\begin{table}
  \caption{Area per lipid (APL) values of POPC with no ions from the Lipid14 simulation
    ran in this work and from the literature, from ECC-lipid model and from experiments.\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 62.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}
